Identifying 15-minute neighbourhoods with complete caring potential

### 2024-09-27 paper/Manuscript.qmd
Still working on the draft...

---
title: "Towards caring 15-minute neighbourhoods"
author:
  - name: Anastasia Soukhov 
    email: soukhoa@mcmaster.ca
    affiliations:
        - name: McMaster University
          department: School of Earth, Environment & Society
          city: Hamilton, Canada
    attributes:
        corresponding: true
  - name: Léa Ravensbergen
    email: ravensbl@mcmaster.ca
    affiliations:
        - name: McMaster University
          department: School of Earth, Environment & Society
          city: Hamilton, Canada
  - name: Lucía Mejía Dorantes
    email: lucia.mejia.dorantes@gmail.com
    affiliations:
        - name: Consultant
          city: Karlsruhe, Germany
  - name: Antonio Páez
    email: paezha@mcmaster.ca
    affiliations:
        - name: McMaster University
          department: School of Earth, Environment & Society
          city: Hamilton, Canada
format: 
  docx
  # elsevier-pdf:
  #   keep-tex: true
  #   include-in-header:
  #     text: |
  #       \usepackage{graphicx}
  #       \usepackage{unicode-math}
  #       \usepackage{hyperref}
  #       \def\tightlist{}
  #       \usepackage{setspace}
  #       \doublespacing
  #       \usepackage{lineno}
  #       \linenumbers
  #   journal:
  #     name: 7th International Conference on Women and Gender in Transportation (WGiT)
  #     formatting: preprint
  #     model: 3p
  #     cite-style: authoryear
keywords: ["Accessibility", "Mobility of Care", "Gender", "15-minute neighbourhood", "Care Complete Neighbourhoods"]
bibliography: bibliography.bib
---

@@

```{r library-setup}
library(biscale)
library(cowplot)
library(magick)
library(dplyr)
library(flextable)
library(ftExtra)
library(glue)
library(here)
library(mapview)
library(pander)
library(scales)
library(sf)
library(tmap)
library(tmaptools)
```

```{r}
load(glue::glue(here::here(), "/data/care_dest.rda"))
```

@@

The "15-Minute Neighbourhood" concept has been adopted by leaders around the world as a way to promote human-scale cities. The concept emphasizes a transport and land-use mix that supports short-trips to daily necessities by foot or bicycle from ones' residence. This study adopts the 15-minute neighbourhood concept through the lens of "Mobility of Care", a framework that foregrounds travel to care destinations, travel done predominately by women, in contrast to the better-studied travel to employment and leisure. While the 15-minute neighbourhood concept is flexible enough to consider all destination types, gendered examinations are lacking in the literature and no research to date has focused explicitly on care destinations. In this study, we address these gaps by identifying which areas in Hamilton, Canada are ‘caring 15-minute neighbourhoods'. To do so, a database of care destinations was compiled and used to estimate both the number (cumulative opportunities indicator) and diversity of mix (entropy measure) of destinations reachable within a 15-minute walk from all residential parcels. Parcel-level results were classified into 'caring' (quantity of reachable opportunities) and 'complete' (diversity of care destinations) neighbourhood typologies of varying degrees using the Self-Organizing Maps (SOM) machine learning technique. Inequalities in the relationship between the identified caring typologies and social-economic indicators were then explored using SOM. Our results suggest the majority of 'caring and complete' neighbourhoods are within the urban core where a higher proportion of lower-income and single-parent households reside. Further, certain neighbourhoods in suburban and rural communities are 'caring and somewhat complete', showing the potential to become 'complete' through possible urban policy intervention. Taken together, this study offers a theoretical bridge to connect Mobility of Care and the 15-minute neighbourhoods concept, a methodology to assess neighbourhoods on a 'caring and complete' continuum, and discussion of relevant gender mainstreaming policy interventions.

@@

The "15-Minute City" has been adopted by leaders as a way to promote human-scaled cities. As defined in @moreno_introducing_2021, the 15-Minute City is a urban planning model based in "chrono-urbanism", a theory that emphasizes the positive impact on quality of urban life when urban space becomes multi-rythmic [related to chronotopy @mulicekUrbanRhythmsChronotopic2015]. Multi-rhythmic urban space in opposition to industrial Fordism and single-use zoning [@mulicekUrbanRhythmsChronotopic2015; @morenoVilleQuartHeure2016]; the segregation of urban time and space for individual uses and mobility. The 15-Minute City is one that accommodates the ability to reach essential destinations such as shops, work, and healthcare within a walkable radius. It would allow individuals to reclaim time spent on car mobility and give way to sustainable modes; prompting urban spaces that are responsive to human needs and environmental sensibilities [@Allam2022]. The 15-Minute City generated much interest in planning practice and public discourse, especially after Paris mayor Anne Hidalgo proposed a 15-minute city plan in her 2020 re-election campaign. Though the term 15-Minute City is relatively new, the core of the concept is not.

Literature has drawn parallels from the 15-Minute City to Perry's debated Neighbourhood Unit Concept (NUC) from the 1930s [@kissfazekasCircleParadigms15minute2022]; a 'socio-spatial' normative scheme widely adopted by government agencies (and the real-estate community) in the Western world after the Second World War [@talenSocialSciencePlanned2017; @solow1969concept]. Pairing well with the objectives of planning agencies at the time, Perry's NUC would allow for efficient mass-building of cellular units that prioritized the perceived functional needs of women and children: each unit providing proximate access to an elementary school and sufficent supporting community facilities, shopping, parks and housing [@talenSocialSciencePlanned2017; @brodyNeighbourhoodUnitConcept2013]. The NUC primarily prioritized local service provision, though Perry had confidence in good design's contribution to 'neighbourhood spirit' [@hall2014cities]. From the NUC's conception until the end of the 1960s, planner's attempts to prescribe social meaning to the neighbourhood's physical form had been criticized to near extinction [@talenSocialSciencePlanned2017]. By social scientists, the overestimation on the impact of built environment on social life was critiqued. By other planners, the inability to agree on the specificity of the neighbourhood (i.e., population sized, and the type, quality, and quantity of amenity) as well as how neighbourhood units are connected to other neighbourhoods and the rest of the region. <!-- failing to combine work, home and cultre was a dilbirate outcome of state-sponsored capitalism X. As stated in [@talenSocialSciencePlanned2017],-->  In response to these criticisms, neighbourhood planning proponents have redefined deterministic terminology: prescribed physical forms is 'enabling' rather than engendering, and the provision of social contact 'opportunities' instead of determined social behaviour. The redefined bottom-up concept to community and neighbourhood planning has been adopted by New Urbanist from which the 15-Minute City stems. The 15-Minute City is created through bottom-up proximity-based processes of community planning [@kissfazekasCircleParadigms15minute2022].

<!-- These critiques were supported by case studies of planned neighbourhoods as... lead to social segregation of classes, destination functions, and automobile use. From it's onset, it was an efficient way to develop housing after the Second World War. The emphasis of proximity to some 'daily essentials' but disconnection from specialised activites (e.g., work) paved the way for car-based subrbia.  by capital of cities. Can the neighbourhood form might be seen as a vehicle for capitalist exploitation, gentrification and displacement??? The place of neighbourhood planning is actively being debated: can it's social crtiques be addressed in the new form as a "15-Minute City"? This study seeks to contribute to this decades old debate. -->

However, critiques of social scientists and planners from decades past still persist. Can physical form be planned to enable an improved quality of life? This line of reasoning has occupied urban planning research in recent years: what are the impacts of built environment on residence's outcomes. A popular tool to analyse this relationship has often been accessibility analysis: proximity to X amenity results in X outcomes for a neighboourhood. Accessibility is a way of quantifying the ease of opportunity interaction from some point in space, this indicator is becoming increasingly important for planners in the pursuit of just and sustainable city planning [@vale2023accessibility; @handyAccessibilityIdeaWhose2020]. As a amenity-provision concept of neighbourhood planning that seeks to enable an improvement on urban quality of life, the 15-Minute City lends well to analysis using accessibility tools.

Before discussing 'what outcome', we first ask 'what amenity' as the focus of this study. Often, accessibility analysis studies have focused on a single amenity type (X), often employment destinations and often focused on a scale beyond the neighbourhood. However, due to 15-Minute City's emphasis on proximate-level service provision, recent academic work on the topic has employed accessibility analysis to account for multiple 'local' amenities types, but seldom included employment. We argue that the inclusion of employment opportunities into the 15-Minute City conceptualisation is critical to it's definition. Why? We echo critics made by planners and social scientistics in decades past: neighbourhoods that did not mix work with service and culture within the neighbourhood, paved the way for car-based suburbia: cars allowed for the connection of neighbourhoods with their 'lower-order' amenities to areas with 'high-order' destinations like work (X). The lower- and higher- order framing of these destinations supported the continued segregation of urban life across many dimensions (X).

Combing all amenity types into the 15-Minute City concept is also a matter of gender equality. There is a branch of research that encompases this critique: Mobility of Care. This framework encourages the explicit consideration of destination types that serve unpaid labour (care destinations) into transport planning (as opposed to only considering destinations with paid work). Not considering these care destinations have travel time (and hence quality of life) implications on those who do care activites, commonly women and especially disadvantaged women. The consideration of _what_ amenity type is a gender issue.

From this position, this study aims to holistically investigate the ease of interacting with proximate amenities (accessibility) from the perspective of the 15-Minute City and the Mobility of Care framework. As mentioned, accessibility analysis have typically focused on employment (X) and recent 15-Minute City flavoured analysis focus on 'lower-order' amentities. We focus on a case study of the mid-sized City of Hamilton, Canada, and aim to classify areas by what degree they may be considered a '15-Minute Neighbourhood" based on the relative quantity of amenity access and diversity of access. We offer a novel methodology of holistically considering all proximate-amenities, offer empirical findings, and contribute theoretically to the ongoing debate on the definition of a 15-Minute City by foregrounding care destinations.

--------------

Caring activities meet the physical, psychological, and emotional needs of others, and are one of the most essential and fundamental activities in society [@ilo_care_2018]. Yet, they are the most unequal, undervalued, and even devalued activities worldwide. Traditionally, caring activities have been borne on women’s shoulders [@haydenGrandDomesticRevolution1982; @hochschild_second_2012]. According to @ilo_care_2018, women and girls perform more than three-quarters of the total amount of unpaid care, a gender gap that varies geographically [@RN5].

This unequal share of caring responsibilities leads to multifaceted gendered differences: in career development, profession selection, contract type, pay gap, and time poverty, as recognized by various international organisations [@EIGE2016; @ilo_care_2018]. In fact, early feminists in the United States had already observed that domestic labor was the single most relevant cause of women’s inequality in society [@gordon_womans_1976; @haydenGrandDomesticRevolution1982].

Care also has important implications for spatial and transportation planning. Indeed, researchers from various fields have highlighted that western modern industrial societies have imposed a physical separation of the private sphere from the public one, splitting production and reproduction, public and private spaces or simply work and home, concentrating on the economic outcome of being mobile [@cresswell_gendered_2008; @lawWomenTransportNew1999; @levy2013travel; @little_gender_1994; @tronto_toward_1990]. Men tend to be associated with legal rights, paid labour, and the public sphere while woman are associated with the private sphere, dependents, family duties, personal relations and unpaid labour [@tronto_toward_1990]. In this line, @cresswell_gendered_2008 described how masculinity has been constructed as mobile and active, whereas femininity is static. As an example of this dichotomy, transport planning has historically framed non-work/school trips as 'non-mandatory' activities, downgrading these activities in comparison to mandatory trips like work and school (XXX). In more recent years, planners have begun to resist the spatial separation of the public and private sphere to promote for sustainable cities. Proximity planning has become more normalized, a popular example is the 15-minute neighbourhood i.e., neighbourhoods where residents can reach all necessary destinations within a 15-minute active transport trip. The 15-minute neighbourhood concept, however, has yet to foreground care.

While much care labour takes place in the private and feminised sphere, completing care activities requires one to be mobile. In other words: caring for oneself and others requires trips to, amongst other places, grocery stores, day cares, elder care, and health care destinations. From this research motivation, Sánchez de Madariaga coined the term ‘mobility of care’ in 2013 to refer to all travel required to sustain the needs of a household, such as grocery shopping, escorting children, travelling to health appointments, and running errands [@sanchezdemadariagaMobilityCareIntroducing2013]. While decades of research had examined these unique household-serving trips, her seminal work was the first to consider all these trips as one category. In doing so, she showed how mobility of care is a significant proportion of daily travel. Indeed, in her original work, and in the studies that have followed, it appears that approximately 30% of adults’ trips are for care purposes [@sanchezdemadariagaMobilityCareIntroducing2013; @sanchezdemadariagaMeasuringMobilitiesCare2019; @ravensbergen2023exploratory]. Perhaps unsurprisingly given the gendered division of care work discussed above, women have consistently been found to complete more mobility of care trips than men. In one study, mobility of care comprised 32% of women’s daily trips compared to 28% of men’s. While this gap was significant, it was found to be far greater in lower income households where women complete 20% more care trips than men [@ravensbergen2023exploratory]. Sánchez de Madariaga not only shows how important these mobility of care trips are, but also highlights the ways in which “mobility of care is systematically under-represented in any analysis of urban transport” (p. 37). Transport surveys and tools do not directly capture mobility of care, which re-enforces the idea that these trips are not a significant part of daily mobility.

Cities have simplified individuals as a universal and neutral entity. As a result, public spaces are differently experienced depending on our multiple identities. Indeed, asymmetric in the distribution of time of citizens is deeply anchored with the public and private spaces organization within cities. In this light, we bring together the literatures on 15-minute cities and mobility of care by focusing on the potential access to destinations of care through an empirical case study of the mid-sized city of Hamilton, Canada. We acknowledge that caring is composed by actors and activities but aim to spotlight the role that the urban environment plays on our daily lives, on easing our participation and on exercising our rights to the city as individuals with multiple identities. Our work hopes to contribute to a fruitful discussion that may trigger urban and transport policy interventions that increase gender-equity in caring activities among all actors.

## Feminist 15-minute neighbourhoods

`r word_comment("TO ADD: some good feminist 15 minute cities. It's novel but movements exists in the Canadian context i.e., https://x.com/brenttoderian/status/1800371515877384527?s=48&t=QI0WSvV_FDb82ZGeLdJ9lQ. Also discuss Barcelona and Bogota (care blocks) non academic sources: https://oecd-opsi.org/innovations/bogota-care-blocks/ AND https://www.npr.org/sections/goatsandsoda/2023/10/16/1205218707/how-bogota-cares-for-its-family-caregivers-from-dance-classes-to-job-training )", " ")`

<!-- The spatial structure of cities and regions has been described in different ways and its advantages and disadvantages has been discussed with regard to a wide variety of issues, for example, regarding health, the environment, economic, social and equity impacts. After the traditional monocentric vision of the city with a core Centre Business District (CBD), other alternatives appeared, especially due to the expansion of cities caused by the automobile and motorized transport. The polycentric urban planning emerged based on the idea of sub-centers (or multiple business centers) with their own services and functions, well-connected among other surrounding areas, where the links between land-use and transport were optimised. Economic activities (such as retail or commercial office development) and job centers played an important role in shaping the spatial structure [@schneider_transit_1981; @sweet_are_2017]. This concept may have different interpretations depending on the spatial and geographical scales [@hall_polycentricity_2009]. -->

In the last decades, the need for more sustainable, healthier, and liveable cities has received more attention. Planners and policy makers have proposed different alternatives where residents reach “essential activities” within walkable or cycling distances, reducing the need for using motorised transportation modes. The 15-minutes cities or "x"-minutes cities, have been long under the spotlight. This approach has also received the name of Chrono-urbanism or the “cities of proximity” [@logan_xminute_2022; @moreno_introducing_2021].

In the international context, there are different examples that highlight the importance of good access to different services and amenities within walkable distances. Probably one of the first cities to adopt the proximity approach was Portland. The Portland Plan [@city_of_portland_20minute_2010] was adopted by the City Council on April 25, 2012 aiming at fostering an inclusive city development based on prosperity, education, health, and equity for the following 30 years. The progress report describes the tasks of the established Office of Equity and Human Rights was established focusing on equitable service delivery to all residents. In the first example, equity is constrained to topics related to racial equity or people with disabilities [@portland_government_portland_2017]. `r word_comment("NOTE TO SELF: And so what? Is this a research gap?", " ")`

Later, other examples have highlighted the benefits of mobility of proximity without specifically targeting mobility of care, where a universal and neutral entity approach persists. In Paris, the adoption of the 15-minute neighbourhood put into the spotlight this approach. They targeted six social functions that had to be easily accessible at any part of the city: housing, work, health care, groceries, education and leisure [@ville_de_paris_paris_2022]. In 2021, Ottawa also adopted a 15-Minutes approach in the new City of Ottawa Official Plan [@ville_dottawa_quartier_2021]. More recently, @moreno_introducing_2021 considers four dimensions on how the 15-minute neighbourhood framework: density (in terms of people per sq.km); diversity (including mixed-land use and diversity of people; the temporal and spatial proximity to essential services (residential areas, work, commercial areas, education centres, health facilities and other basic institutions) and digitalisation.  In this respect, the feminist perspective of the cities of proximity is still underestimated. Only few examples exist that specifically target this issue [@gil_sola_choose_2022; @macintyre_her_2022]. Given the impact of caring activities on everyday life, the chrono-urbanism should specifically target the analysis of caring activities.

Within the backdrop of urban planning and ‘mobility of care’, we connect our work to the 15-minute neighbourhood concept.

>>>>

---
title: "Identifying 15-minute neighbourhoods with complete caring potential"
author:
  - name: Anastasia Soukhov 
    email: soukhoa@mcmaster.ca
    affiliations:
        - name: McMaster University
          department: School of Earth, Environment & Society
          city: Hamilton, Canada
    attributes:
        corresponding: true
  - name: Léa Ravensbergen
    email: ravensbl@mcmaster.ca
    affiliations:
        - name: McMaster University
          department: School of Earth, Environment & Society
          city: Hamilton, Canada
  - name: Lucía Mejía Dorantes
    email: lucia.mejia.dorantes@gmail.com
    affiliations:
        - name: Consultant
          city: Karlsruhe, Germany
  - name: Antonio Páez
    email: paezha@mcmaster.ca
    affiliations:
        - name: McMaster University
          department: School of Earth, Environment & Society
          city: Hamilton, Canada
format: 
  docx
  # elsevier-pdf:
  #   keep-tex: true
  #   include-in-header:
  #     text: |
  #       \usepackage{graphicx}
  #       \usepackage{unicode-math}
  #       \usepackage{hyperref}
  #       \def\tightlist{}
  #       \usepackage{setspace}
  #       \doublespacing
  #       \usepackage{lineno}
  #       \linenumbers
  #   journal:
  #     name: 7th International Conference on Women and Gender in Transportation (WGiT)
  #     formatting: preprint
  #     model: 3p
  #     cite-style: authoryear
keywords: ["Accessibility", "Mobility of Care", "Gender", "15-minute neighbourhood", "Care Complete Neighbourhoods"]
bibliography: bibliography.bib
---

@@

```{r library-setup}
library(biscale)
library(cowplot)
library(magick)
library(dplyr)
library(flextable)
library(ftExtra)
library(glue)
library(here)
library(mapview)
library(pander)
library(scales)
library(sf)
library(tmap)
library(tmaptools)
library(gt)
```

```{r}
load(glue::glue(here::here(), "/data/care_dest.rda"))
load(glue::glue(here::here(), "/data/access_15_wide.RData"))
```

@@

The "15-Minute Neighbourhood" concept has been adopted by leaders around the world as a way to promote human-scale cities. The concept emphasizes a transport and land-use mix that supports short-trips to daily necessities by foot or bicycle from ones' residence. This study adopts the 15-minute neighbourhood concept through the lens of "Mobility of Care", a framework that foregrounds travel to care destinations, travel done predominately by women, in contrast to the better-studied travel to employment and leisure. While the 15-minute neighbourhood concept is flexible enough to consider all destination types, gendered examinations are relatively lacking in the literature and no research to date has focused explicitly on care destinations. In this study, we address these gaps by identifying which areas in Hamilton, Canada have the potential to be ‘caring 15-minute neighbourhoods'. To do so, a database of care destinations was compiled and used to estimate both the number (cumulative opportunities indicator) and diversity of mix (entropy measure) of destinations reachable within a 15-minute walk from all residential parcels. Parcel-level results were classified into 'caring' (quantity of reachable opportunities) and 'complete' (diversity of care destinations) neighbourhood typologies of varying degrees using machine learning techniques (Self-Organizing Maps (SOM) and hierarchical clustering). Inequalities in the relationship between the identified caring typologies and social-economic indicators of who currently reside in the neighbourhoods were then explored. Our results suggest the majority of 'caring and complete' neighbourhoods are within the urban core where a higher proportion of lower-income and single-parent households reside. Further, certain neighbourhoods in suburban and rural communities are 'caring and somewhat complete', showing the potential to become 'complete' through possible urban policy intervention. Taken together, this study offers a theoretical bridge to connect Mobility of Care and the 15-minute neighbourhoods concept, a data-driven classification and clustering methodology to assess neighbourhoods on a 'caring and complete' continuum, and discussion of relevant gender mainstreaming policy interventions.

@@

The "15-Minute City" has been recently adopted by leaders as a way to promote human-scaled cities (e.g., Paris mayor Anne Hidalgo proposed a 15-minute city plan in her 2020 re-election campaign, 15-minute walkable neighbourhood guidance in Shanghai, China [@weng15minuteWalkableNeighborhoods2019]), As defined in @moreno_introducing_2021, the 15-Minute City is a urban planning model based in chrono-urbanism, a theory that emphasizes the positive impact on quality of urban life when urban space becomes multi-rythmic, in opposition to industrial Fordism, single-use zoning, and the segregation of urban time and space for individual uses and mobility [@mulicekUrbanRhythmsChronotopic2015; @morenoVilleQuartHeure2016]. The 15-Minute City is one that accommodates the ability to reach essential destinations within a walkable, bikeable or transit supportive radius. It would allow individuals to reclaim time spent on car mobility and give way to sustainable modes; prompting urban spaces that are responsive to human needs and environmental sensibilities [@Allam2022].

Though the term 15-Minute City is relatively new, neighbourhood planning practice aiming to foster community and local travel to amenities is not. As recent examples, planned neighbourhoods dominated post-WWII built urban form: the Neighbourhood Unit Concept in the western world [@brodyNeighbourhoodUnitConcept2013] and a parallel concept of the “mikrorayons” neighbourhoods in the Soviet Union  [@kissfazekasCircleParadigms15minute2022]. These planning forms have also been extensively critiqued: as summarized in @talenSocialSciencePlanned2017, urban planners struggled to determine suitably universal design aspects such as neighborhood size, the type and quality of amenities, and local and regional connectivity. Social scientists also emphasized that top-down neighbourhood planning approaches overestimated the built environment's influence on social life [@talenSocialSciencePlanned2017]. Neighbourhood planning approaches such as the 15-Minute City concept are being re-embraced, but from bottom-up perspective. However how to equitably prescribe urban form that 'enables' contact with social opportunities (instead of 'engendering it' from the top-down) is still under debate. For instance, questions such as _what_ are destinations that matter, by what mode and travel time threshold, and _for whom_ remain under discussion among proximity-based planners [@silvaProximitycentredAccessibilityConceptual2023; @silvaRegionalAccessibilityUndermining2022].

To add clarity to these questions surrounding the 15-Minute City concept, in this work we aim to build a theoretical bridge to "Mobility of Care" as introduced in @sanchezdemadariagaMobilityCareIntroducing2013. Mobility of Care is especially relevant to the 15-Minute City, as it is a purpose of travel that is necessary for all and often occurs at the local level. Namely, Mobility of Care refers to all travel required to sustain the needs of a household, such as grocery shopping, escorting children, travelling to health appointments, and running errands. While decades of research have examined types of household-serving trips (e.g., shopping trips, escorting to school), @sanchezdemadariagaMobilityCareIntroducing2013 was the first to consider them all as one category. In doing so, Mobility of Care is shown to be a significant proportion of daily travel, characterizing the trip purpose of approximately 30% of adults’ trips [@sanchezdemadariagaMobilityCareIntroducing2013; @sanchezdemadariagaMeasuringMobilitiesCare2019; @ravensbergen2023exploratory]. Notably, care trips take on tremendous gendered and socio-economic importance, as women, and especially those from lower income households, complete the majority of care trips [@ravensbergen2023exploratory]. Despite its importance for all and in advancing gender equality, the examination of mobility of care and access to care destinations have been under-examined relative to employment destinations. This is especially pertinent in accessibility research that has largely focused on employment () and retail ().

We aim to theoretically bridge Mobility of Care with the 15-Minute City concept to clarify questions surrounding _what destinations within a 15 minute walkable radius matter_. We hypothesis that _care_ destinations matter, they matter for all people but especially for women and lower-income women preforming care trips. In this vein, this work pairs this hypothesis with an exploratory clustering methodology to identify which neighbourhoods have the potential to support 15-minute access to care destinations in an empirical case study of Hamilton, Ontario, typical mid-sized Canadian city. More concretely, the research objectives are two-fold: after a review of the 15-Minute City concept and associated measurement tools, this work first introduces the Mobility of Care conceptualisation to re-imagine local amenities importance within the 15-Minute City concept, and second to apply a data-driven classification and clustering machine learning methodology to support the bottom-up examination of _15-minute caring neighbourhoods_  through an empirical examination of Hamilton.

# Review of neighbourhood planning literature

## From the 15-Minute City to the NUC

In the last decades, the need for more sustainable, healthier, and liveable cities have received more attention. Planners and decision-makers have proposed alternatives to mobility-based neighbourhood planning to one centered on proximity to urban functions [@pozoukidou15MinuteCityDecomposing2021]. In this context, the 15-Minute City is now in the public spotlight [@logan_xminute_2022; @moreno_introducing_2021]. In @moreno_introducing_2021 where the 15-Minute City concept was introduced, four universal dimensions are outlined: density (in terms of people per km$^2$), diversity (including mixed-land use and diversity of people), the temporal and spatial proximity to essential services, <!-- (residential areas, work, commercial areas, education centres, health facilities and other basic institutions)--> and digitalisation. The framework presented has been critised within academic and planning circles [@guzmanProximityEnoughCritical2024]. Namely, critiques have centered on the concept's inadequecy in addressing pre-existing structural forces and individual charctertes that drive inequalities in who does and who could benefit more from such a paradigm [@dimarino15minuteCityConcept2023; @willberg15minuteCityAll2023]. A 15-Minute City for all people is a utopic ideal, however how will existing inequalities in how people move through and benefit from the city be addressed. _Who is in the most need of the 15-Minute City before we all get there?_. Without directed answers to these context-sensitive questions makes this increasingly popular concept susceptible to becoming a place-branding slogan [@pozoukidou15MinuteCityDecomposing2021].

Reflecting this flexible framework; cities that have adopted 15-Minute City concepts have done so using a diverse range of definitions and tools along with similar indistinct universal approaches. An early adopter was Portland, US in the Portland Plan [@city_of_portland_20minute_2010] of April 2012. It aims to foster an inclusive city development based on prosperity, education, health, and equity for a 30 year horizon. It promotes neighbourhood self-suffiency and connectivity to city centers and centres of employment. The progress report describes a high-level focus on equitable service delivery to all residents with equity concerning topics related to racial equity and people with disabilities [@portland_government_portland_2017]; similarly adopting an "all populations" approach. Later, other cities also adopted proximity-based goals into their plans, taking similar neutral entity approaches. For instance, in the plan that later catapulted the 15-Minute City into public discourse: Paris mayor Anne Hidalgo proposed a 15-minute city plan in her 2020 re-election campaign that foregrounded six social functions of the city that should be easily accessible from any location, namely: housing, work, health care, groceries, education and leisure [@ville_de_paris_paris_2022]. These ideas inspired language in the agendas of other cities in the Western world such as Ottawa, in the Canadian context, who also adopted a 15-Minutes approach in their recent Official Plan [@ville_dottawa_quartier_2021]. From the review of worldwide practice literature in @teixeiraClassifying15minuteCities2024, overall the 15-Minute City concept is in early phases of implementation around the world and the diverse range of definitions, strategies, and instruments present a significant knowledge and implementation gap: the 15-Minute City seems nice but _how do we get there and whom will benefit_?

The past can shed light on cautionary tales. While the 15-Minute City concept is new, neighbourhood planning to improve society outcomes is not: in fact the literature has drawn parallels from the 15-Minute City to Clarence A. Perry's Neighbourhood Unit Concept (NUC) from the 1930s [@kissfazekasCircleParadigms15minute2022]. The NUC is a socio-spatial normative scheme widely adopted by government agencies (and the real-estate community) in the Western world after the Second World War [@talenSocialSciencePlanned2017; @solow1969concept]. Pairing well with the objectives of planning agencies at the time, Perry's NUC would allow for efficient mass-building of cellular units that prioritized the perceived functional needs of women and children: each unit providing proximate access to an elementary school and sufficent supporting community facilities, shopping, parks and housing [@talenSocialSciencePlanned2017; @brodyNeighbourhoodUnitConcept2013]. The NUC primarily prioritized local service provision, though Perry had confidence in good design's contribution to 'neighbourhood spirit' [@hall2014cities]. From the NUC's conception until the end of the 1960s, planner's attempts to prescribe social meaning to the neighbourhood's physical form had been criticized to near extinction [@talenSocialSciencePlanned2017]. By social scientists, the overestimation on the impact of built environment on social life was critiqued. By other planners, the inability to agree on the specificity of the neighbourhood (i.e., population sized, and the type, quality, and quantity of amenity) as well as how neighbourhood units are connected to other neighbourhoods and the rest of the region.  In response to these criticisms, neighbourhood planning proponents have redefined deterministic terminology: prescribed physical forms as 'enabling' rather than engendering, and the provision of social contact 'opportunities' instead of determined social behaviour. The re-defined bottom-up approach to community and neighbourhood planning has been adopted by New Urbanist from which the 15-Minute City stems [@kissfazekasCircleParadigms15minute2022]. But with _what bottom-up processes?_; small scale experiments are being tested () with preliminary small-scacle results. These processes are only in early stages so critiques of social scientists and planners from decades past still persist.

_How can physical form be planned to enable an improved quality of life, and for whom?_  has also occupied urban planning research in recent years, exploring what are the impacts of built environment on residence's outcomes. For instance, the examination of low-income households access to transportation and their likelihood of gaining employment [@blumenbergDriveWorkRelationship2017; @bastiaanssenDoesBetterJob2022] or the relationship between children's access to public transit and participation in after-school activities [@palmRolePublicTransit2020]. Relatedly, a new wave of researchers and practitioners focused on local and context-specific relationships in the proximity to destinations have also emerged [@silvaProximitycentredAccessibilityConceptual2023; @silvaRegionalAccessibilityUndermining2022]. As reviewed  in the city plans that have adopted 15-Minute City approaches, the NUC's critisms, and recent proximity-based examinations, the question of _how to enable improved quality of life through urban built environment_ has shown to be highly context-senstive with the need for further investigation.

## Tools: accessibility methods, diversity measures and typology-classification

In examining how to enable improved quality of life through urban built environment, an increasingly popular tool have been accessibility measures. They are a way to quantify the ease of reaching destinations from a given point in space and have been used to examine urban areas through just and sustainable city planning agendas [@vale2023accessibility; @handyAccessibilityIdeaWhose2020]. As the 15-Minute City is an amenity-provision neighbourhood planning concept which aims to enable urban environment that improves life quality, its analysis lends well to the use of accessibility tools [@guzmanProximityEnoughCritical2024]. Foreseeably, recent works have applied accessibility measures to investigate the 15-Minute City across various geographic scopes. For instance, in Naples, Italy [@gaglioneUrbanAccessibility15minute2022], Barcelona, Spain [@graellsCityCitiesMeasuring2021], Vancouver, Canada [@hosford15minuteCityReach2022], and in urban areas across Europe [@vale2023accessibility]. The "cumulative opportunity" measure has been applied to many 15-Minute City accessibility examinations. This form of accessibility quantifies how many destinations can be reached from a point in space within a given travel time threshold, pairing well with examinations that have a normative travel time threshold, like the 15-Minute City. Furthermore, the cumulative opportunity measure is widely appreciated for its intuitive computation and is popular among practitioners [@handyAccessibilityIdeaWhose2020; @handyMeasuringAccessibilityExploration1997; @chengInvestigatingWalkingAccessibility2019]. However, reflecting the diversity of the accessibility literature, accessibility measures other than the cumulative opportunity have also been applied [@guzmanProximityEnoughCritical2024].

Another method which compliments the assessment of the 15-Minute City as it relates to the diversity of opportunity types, is entropy. The entropy measure has been widely used to characterise the diversity of land-use mix in early work by @frankLinkingObjectivelyMeasured2005 based on the species diversity Shannon-Wiener index that expresses relative evenness throughout a sequence (this measure among other species diversity indices of the time are reviewed in @whittaker1972evolution). It has also been used in social sciences to comprehend the occurrence of social events in a structured manner specially with regard to life trajectories [@ritschardSequenceAnalysisRelated2018].In the realm of urban studies and planning literature the malleable concept of entropy has been used to characterise the diversity in land-use mix [@ewingTravelBuiltEnvironment2010a], especially in the context of active modes [@luUrbanDensityDiversity2017; @mavoaIdentifyingAppropriateLanduse2018] and suburban sprawl [@randallGISBasedLand2015]. As well, it has been used to understand mobility behaviour [@mcbride2020exploration; @montero2023applying; @montero2023role]. Work that has used diversity indices alongside accessibility analysis are scarce, though examples of its use as parameters within accessibility scores in the case of employment opportunities [@chengMeasuringUrbanJob2013; @daiIncorporatingJobDiversity2018] and more recently to describe the diversity in transit facilities [@yinIncorporatingFacilityDiversity2024] are present. To the authors knowledge, the use of an entropy measure to reflect the diversity in care destination amenity types and in context within a 15-minute neighbourhood, has yet to be done. However, entropy measures thematically compliments the 15-Minute City.

`r word_comment(Maybe Antonio to add? I need to fill this out a bit", " ")`
In addition, a useful tool has been classification algorithms in the identification of spatial typologies and in profiling who may reside there. Those too have K means... Hierarchical... and SOM [@delmelleDifferentiatingPathwaysNeighborhood2017]... regionlization techniques such as spatially constrained multivariate clustering [@burkeGeospatialAnalysisFramework2022] Use of tree diagrams!

SOM is a type of artificial neural network (ANN) trained using unsupervised learning that produces a two-dimensional representation of the input space. It applies competitive learning by way of a neighbourhood function thereby preserving the topological properties of the input space and producing combinations of the input space that we can then intuitively profile. It has been used widely in XX but scarcely in transportation related research [@victorianoTimeSpaceMoney2020]

All-in-all, within the context of examining the 15-Minute City there has been a wide range of tools applied in the literature that appear appriorpiate. However, the question of _15-minute cities for whom_ requires further examination. In this way, this work connects the 15-Minute City to another normative conceptualisation: Mobility of Care.

## Mobility of Care and feminist 15-minute neighbourhoods

As the 15-Minute City is in early stages of conceptualisation; an alternative way to conceive of _what_ urban functions matter could be through a feminist lens. Instead of focusing on _all_ destinations, it may be useful to examine destinations involved in caring activities. Caring activities meet the physical, psychological, and emotional needs of others, and are one of the most essential and fundamental activities in society [@ilo_care_2018]. Yet, they are the most unequal, undervalued, and even devalued activities worldwide. Traditionally, caring activities have been borne on women’s shoulders [@haydenGrandDomesticRevolution1982; @hochschild_second_2012]. According to @ilo_care_2018, women and girls perform more than three-quarters of the total amount of unpaid care, a gender gap that varies geographically [@RN5]. This unequal share of caring responsibilities leads to multifaceted gendered differences: in career development, profession selection, contract type, pay gap, and time poverty, as recognized by various international organisations [@EIGE2016; @ilo_care_2018]. In terms of spatial and transportation planning, almost one third of daily trips are for care purposes [@sanchezdemadariagaMobilityCareIntroducing2013; @sanchezdemadariagaMeasuringMobilitiesCare2019; @ravensbergen2023exploratory]. From this research motivation, Sánchez de Madariaga coined the term Mobility of Care in 2013 to refer to all travel required to sustain the needs of a household, such as grocery shopping, escorting children, travelling to health appointments, and running errands [@sanchezdemadariagaMobilityCareIntroducing2013]. While decades of research had examined these unique household-serving trips, her seminal work was the first to consider all these trips as one category and demonstrate how mobility of care is a significant proportion of daily travel.
